---
title: "Visualising Covid-19 v2"
output: pdf_document
date: "2023-07-13"
---
```{r}
# Load the readr, ggplot2, and dplyr packages
library(readr)
library(ggplot2)
library(dplyr)

# Read datasets/confirmed_cases_worldwide.csv into confirmed_cases_worldwide
confirmed_cases_worldwide <- read_csv("Data/confirmed_cases_worldwide.csv")

# See the result
confirmed_cases_worldwide
```

Confirmed cases throughout the world:
The table above shows the cumulative confirmed cases of COVID-19 worldwide by 
date. Just reading numbers in a table makes it hard to get a sense of the scale 
and growth of the outbreak. Let's draw a line plot to visualize the confirmed 
cases worldwide.
```{r}
# Draw a line plot of cumulative cases vs. date
# Label the y-axis and x-axis
ggplot(confirmed_cases_worldwide, aes(x = date, y = cum_cases)) +
  geom_line() +
  labs(y = "Cumulative Confirmed Cases",
       x = "Date")
```
China compared to the rest of the world:
The y-axis in that plot is pretty scary, with the total number of confirmed 
cases around the world approaching 200,000. Beyond that, some weird things are 
happening: there is an odd jump in mid February, then the rate of new cases 
slows down for a while, then speeds up again in March. We need to dig deeper to 
see what is happening.

Early on in the outbreak, the COVID-19 cases were primarily centered in China. 
Let's plot confirmed COVID-19 cases in China and the rest of the world 
separately to see if it gives us any insight.
```{r}
# Read in datasets/confirmed_cases_china_vs_world.csv
confirmed_cases_china_vs_world <- 
  read_csv("Data/confirmed_cases_china_vs_world.csv")

# See the result
glimpse(confirmed_cases_china_vs_world)

# Draw a line plot of cumulative cases vs. date, colored by is_china
# Define aesthetics within the line geom
plt_cum_confirmed_cases_china_vs_world <- 
  ggplot(confirmed_cases_china_vs_world,
         aes(x = date, y = cum_cases, color = is_china)) +
  geom_line() +
  labs(y = "Cumulative Confirmed Cases",
       x = "Date",
       colour = "Location:")
  
# See the plot
plt_cum_confirmed_cases_china_vs_world
```
In February, the majority of cases were in China. That changed in March when 
it really became a global outbreak: around March 14, the total number of cases 
outside China overtook the cases inside China. This was days after the 
WHO declared a pandemic.

There were a couple of other landmark events that happened during the outbreak. 
For example, the huge jump in the China line on February 13, 2020 wasn't just a 
bad day regarding the outbreak; China changed the way it reported figures on 
that day (CT scans were accepted as evidence for COVID-19, rather than only 
lab tests).

By annotating events like this, we can better interpret changes in the plot.
```{r}
who_events <- tribble(
  ~ date, ~ event,
  "2020-01-30", "Global health\nemergency declared",
  "2020-03-11", "Pandemic\ndeclared",
  "2020-02-13", "China reporting\nchange"
) %>%
  mutate(date = as.Date(date))

who_events

# Using who_events, add vertical dashed lines with an xintercept at date
# and text at date, labeled by event, and at 100000 on the y-axis
plt_cum_confirmed_cases_china_vs_world +
  geom_vline(data = who_events, mapping = aes(xintercept = date), 
             linetype = "dashed") +
  geom_text(data = who_events, mapping = aes(x = date, y = 100000, 
                                             label = event)) +
  labs(y = "Cumulative Confirmed Cases",
       x = "Date",
       colour = "Location:")
```
Adding a trend line to China:
When trying to assess how big future problems are going to be, we need a 
measure of how fast the number of cases is growing. A good starting point is to 
see if the cases are growing faster or slower than linearly.

There is a clear surge of cases around February 13, 2020, with the reporting 
change in China. However, a couple of days after, the growth of cases in China 
slows down. How can we describe COVID-19's growth in China after February 15, 
2020?
```{r}
# Filter for China, from Feb 15
china_after_feb15 <- confirmed_cases_china_vs_world %>%
  filter(is_china == "China", date >= "2020-02-15")

# Using china_after_feb15, draw a line plot cum_cases vs. date
# Add a smooth trend line using linear regression, no error bars
ggplot(china_after_feb15, 
       aes(x = date, y = cum_cases)) +
  geom_line() +
  stat_smooth(method = "lm", se = FALSE) +
  labs(y = "Cumulative Confirmed Cases",
       x = "Date")
```
And the rest of the world?
From the plot above, the growth rate in China is slower than linear. That's 
great news because it indicates China has at least somewhat contained the virus 
in late February and early March.

How does the rest of the world compare to linear growth?
```{r}
# Filter confirmed_cases_china_vs_world for not China
not_china <- confirmed_cases_china_vs_world %>% 
  filter(is_china == "Not China")

# Using not_china, draw a line plot cum_cases vs. date
# Add a smooth trend line using linear regression, no error bars
plt_not_china_trend_lin <- ggplot(not_china, 
                                  aes(x = date, y = cum_cases)) +
  geom_line() +
  stat_smooth(method = "lm", se = FALSE) +
  labs(y = "Cumulative Confirmed Cases",
       x = "Date")

# See the result
plt_not_china_trend_lin 
```
Adding a logarithmic scale:
From the plot above, we can see a straight line does not fit well at all, 
and the rest of the world is growing much faster than linearly. 
What if we added a logarithmic scale to the y-axis?
```{r}
# Modify the plot to use a logarithmic scale on the y-axis
plt_not_china_trend_lin + 
  scale_y_log10()
```
Which countries outside of China have been hit hardest?
With the logarithmic scale, we get a much closer fit to the data. 
From a data science point of view, a good fit is great news. Unfortunately, 
from a public health point of view, that means that cases of COVID-19 in the 
rest of the world are growing at an exponential rate, which is terrible news.

Not all countries are being affected by COVID-19 equally, and it would be 
helpful to know where in the world the problems are greatest. Let's find the 
countries outside of China with the most confirmed cases in our dataset.
```{r}
# Get the data for each country
confirmed_cases_by_country <- read_csv("Data/confirmed_cases_by_country.csv")
glimpse(confirmed_cases_by_country)

# Group by country, summarize to calculate total cases, find the top 7
top_countries_by_total_cases <- confirmed_cases_by_country %>%
  group_by(country) %>%
  summarise(total_cases = max(cum_cases)) %>%
  top_n(7)

# See the result
top_countries_by_total_cases
```
Plotting hardest hit countries as of Mid-March 2020
Even though the outbreak was first identified in China, there is only one 
country from East Asia (South Korea) in the above table. Four of the listed 
countries (France, Germany, Italy, and Spain) are in Europe and share borders. 
To get more context, we can plot these countries' confirmed cases over time.
```{r}
# Read in the dataset confirmed_cases_top7_outside_china.csv
confirmed_cases_top7_outside_china <- 
  read_csv("Data/confirmed_cases_top7_outside_china.csv")

# Glimpse at the contents of confirmed_cases_top7_outside_china
glimpse(confirmed_cases_top7_outside_china)

# Using confirmed_cases_top7_outside_china, draw a line plot of
# cum_cases vs. date, colored by country
top_7_graph <- ggplot(confirmed_cases_top7_outside_china,
       aes(x = date, y = cum_cases, colour = country)) +
  geom_line() +
  theme_bw() +
   labs(y = "Cumulative Confirmed Cases",
       x = "Date",
       colour = "Country:",
       title = "Covid-19 testing in South Korea curbs the rise in cases") +
  theme(plot.title = element_text(hjust = 0.5))

top_7_graph
```












